Exponential Integrators for PDEs arising in meteorology

### 2024-08-31 article_4.tex
Added Kyrlov Subspace Results

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\begin{document}

@@

\begin{algorithm}
\caption{Arnoldi \cite{Fan2018}} %find better citation
\begin{algorithmic}
\Procedure{Arnoldi}{$A, \hat v_1,m$}
\State $v_0 = 0$
\For{$j = 1,2,...,m$}	
\For{$i = 1,2,...,j$}
\State$h_{ij} \gets v_i^T A v_i$
\EndFor
\State$\theta_j \gets Av_j - \sum^j_{i=1} h_{ij}v_i$
\State$h_{j+1,j} \gets ||\theta_j||$
\State$v_{j+1} \gets \theta_j/h_{j+1,j}$
\EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}
Here V is given by $v_1,...,v_m$ and H is give by $h_1,...,h_m$
\begin{algorithm}
\caption{Lanczos \cite{OJALVO1970}}
\begin{algorithmic}
\Procedure{Lanczos}{$A, \hat v_1,m$}
\State $v_0 = 0$
\For{$i \in \{1,2,...,m\}$}	
\State$\beta_i \gets || \hat v_i ||$
\State$v_i \gets \hat v_i / || \hat v_i ||$
\State$\alpha_i = v_i^T A v_i$
\State$\hat v_{i+1} = Av_i - \alpha_iv_i - \beta_iv_{i-1}$
\EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}\\
Here $V$ is given by ${v_1,...,v_m}$ and $H$ is tridiagonal with the leading diagonal being $\alpha_1, ..., \alpha_m$ and the upper and lower diagonals being $\beta_2,...,\beta_m$.
\newpage
\bibliographystyle{plain}
\bibliography{References.bib}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{graphicx}
\usepackage{afterpage}
\begin{document}

@@

\begin{algorithm}
\caption{Arnoldi \cite{Fan2018}} %find better citation
\begin{algorithmic}
\Procedure{Arnoldi}{$A, \hat v_1,m$}
\State $v_0 = 0$
\For{$j = 1,2,...,m$}	
\For{$i = 1,2,...,j$}
\State$h_{ij} \gets v_i^T A v_i$
\EndFor
\State$\theta_j \gets Av_j - \sum^j_{i=1} h_{ij}v_i$
\State$h_{j+1,j} \gets ||\theta_j||$
\State$v_{j+1} \gets \theta_j/h_{j+1,j}$
\EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}
Here V is given by $v_1,...,v_m$ and H is give by $h_1,...,h_m$.\\
\begin{algorithm}
\caption{Lanczos \cite{OJALVO1970}}
\begin{algorithmic}
\Procedure{Lanczos}{$A$ symetric$, \hat v_1,m$}
\State $v_0 = 0$
\For{$i \in \{1,2,...,m\}$}	
\State$\beta_i \gets || \hat v_i ||$
\State$v_i \gets \hat v_i / || \hat v_i ||$
\State$\alpha_i = v_i^T A v_i$
\State$\hat v_{i+1} = Av_i - \alpha_iv_i - \beta_iv_{i-1}$
\EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}\\
Here $V$ is given by ${v_1,...,v_m}$ and $H$ is tridiagonal with the leading diagonal being $\alpha_1, ..., \alpha_m$ and the upper and lower diagonals being $\beta_2,...,\beta_m$.

\subsection{Relevance To Matrix Exponentials}
By using $A \approx VHV^T$ and the fact that $V^TV = I$ we can apply this to $e^A$ as follows:
\begin{align*}
e^A &= \sum^{\infty}_{i=0}\frac{A^i}{i\!}\\
&= \sum^{\infty}_{i=0}\frac{(VHV^T)^i}{i\!} \\
&= \sum^{\infty}_{i=0}\frac{VH^iV^T}{i\!} \\
&\text {and then when computing $e^Av$ we get}\\
e^Av &= (\sum^{\infty}_{i=0}\frac{VH^iV^T}{i\!})v \\
&= \sum^{\infty}_{i=0}\frac{VH^iV^Tv}{i\!} \\
&= \sum^{\infty}_{i=0}\frac{VH^ie_1}{i\!} \\
&= V(\sum^{\infty}_{i=0}\frac{H^i}{i\!})e_1 \\
&= Ve^He_1
\end{align*}

\subsection{Numerical Results}
Here we run numerical experiments, computing $e^Av$ for a tridiagonal matrix $A$ of sizes $100,000$ and $1,000,000$. We will compare the methods with respect to computation time, error and size of the Kyrlov subspace.

\begin{figure}

    \centering
    \begin{minipage}{0.5\textwidth}
       \centering
	  \includegraphics[width=\linewidth]{Plots/M v E Results for N=1000000.png}
	  \label{fig:MEe7}
       \centering
	  \includegraphics[width=\linewidth]{Plots/M v Comp Time Results for N=1000000.png}
	  \label{fig:MEe7}
       \centering
	  \includegraphics[width=\linewidth]{Plots/Comp Time v E Results for N=1000000.png}
	  \label{fig:MEe7}
    \end{minipage}\hfill
    \begin{minipage}{0.5\textwidth}
       \centering
	  \includegraphics[width=\linewidth]{Plots/M v E Results for N=100000.png}
	  \label{fig:MEe6}
       \centering
	  \includegraphics[width=\linewidth]{Plots/M v Comp Time Results for N=100000.png}
	  \label{fig:MEe6}
       \centering
	  \includegraphics[width=\linewidth]{Plots/Comp Time v E Results for N=100000.png}
	  \label{fig:MEe6}
    \end{minipage}
These results show that the Lanzcos algorithm outperforms the Arnoldi algorithm in both convergence rate and computational time.
\end{figure}

\begin{figure}
Below we compare the time required for the different algorithms to get below an error of $10^{10}$ when computing $e^Av$ for various matrix sizes.
\includegraphics[width=\linewidth]{Plots/time to get below an error of 1e-10.png}
\clearpage
Again we continue to see that the Lanzcos algorithm outperforms the Arnodli algorithm.
\end{figure}
\clearpage
\newpage
\bibliographystyle{plain}
\bibliography{References.bib}
